Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-05 manuscript_v2.tex
updated

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

@@

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header
\usepackage{floatrow}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{float}
\floatplacement{figure}{H}
\floatplacement{table}{H}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{times}
\usepackage{smartdiagram}
\usepackage{ragged2e}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{./sty/langsci-gb4e}
\usepackage{caption}
\usepackage{multicol}
\setlength{\columnsep}{-3cm}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Reference Tracking Strategies of Native and Late Signers in
Turkish Sign Language}

  \begin{abstract}
  We investigated the reference tracking strategies among native and
  late adult deaf signers of Turkish Sign Language. Consistent with the
  theories of saliency and referential accessibility, regardless of
  their acquisiton groups, signers mainly used nominals and extension
  classifiers for introducing referents. To maintain a referent in the
  previous clause, signers chose null marking (e.g., constructed action,
  agreement and plain verbs). For topic shifts or re-introduced
  contexts, nominals and pronouns were chiefly favored although we
  observed very little pronominal use. As for age of acquisition
  effects, we only report limited over-redundancy for late signers who
  used null marking less compared to native signers especially for
  introduced and maintained contexts. Given our limited support for
  redundancy by late signers, we conclude that native signers might
  utilize the spatial affordances of the visual-spatial modality better
  than late signers only to a certain extent.
  \end{abstract}
   \begin{keyword} Referential accessibility, signed narratives,
pragmatic competence, delayed language acquisition, Turkish Sign
Language\end{keyword}
 \end{frontmatter}

@@

Narratives are extensive discourses that construct an important part of
human communication. In order to create a systematic and coherent
organization of experiences, narrators must choose appropriate
linguistic forms to refer to entities (Ariel, 1990; Givon, 1983; Gundel
et al., 1993): Certain forms such as nominal expressions are commonly
used for the first mention of a referent in a discourse, and other forms
such as pronouns and null marking may track previously mentioned
entities which could either be maintained across multiple clauses or
re-introduced back into the discourse. The selection of these referring
expressions (henceforth REs) is contingent on cognitive abilities, such
that individuals must rely on the activation of the working memory to
keep in mind previously referred expressions and long strings of
information (Bamberg, 1997; Morgan, 2005). Pragmatic abilities too play
a pivotal role in the tracking of referents. The form with the least
amount of redundant information must be selected based on the
accessibility levels of the referent (i.e., how active a referent is in
the addressee's mind given a point in discourse) (Ahn, 2019; Ariel,
1990). Because of these cognitive and pragmatic necessities, it is not
surprising to observe that producing a coherent narrative is one of the
latest milestones in language acquisition among children (Hickmann et
al., 1996). As can be expected, there is a plethora of research
examining how narrative competence is achieved in both monolingual and
bilingual acquisition contexts among children and adults. Although
narrative competence and referent selection have been studied
extensively for typically developing speakers (Aksu-Koç and
Nicolopoulou, 2015; Hickmann et al., 1996; Williams, 1988), deaf
individuals who acquired a sign language in atypical learning contexts
(i.e., deaf signers with linguistic deprivation) have remained mostly
understudied with an exception of handful of studies (Becker, 2009;
Cormier et al., 2013; Gür, 2018). Nevertheless, it is paramount that we
investigate and describe the effects of language deprivation among deaf
adults on narrative competence more in depth. This will help us
understand whether the loss of systematic and frequent language input in
the early years of life affect how deaf signers produce a narrative in
their adulthood. Given this, the present study will focus on deaf adult
signers with language deprivation and examine their reference tracking
strategies in signed narratives.

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 (see Nuhbalaoglu, 2018 for a brief review of the historical
development of the language). Deaf or hearing children acquiring a SL
within a family where there is one or more deaf caregivers display the
same developmental milestones as hearing children acquiring a spoken
language in a hearing family (Chamberlain et al., 1999; Meier and
Newport, 1990). However, reports indicate that over 90\% of all deaf
individuals are born into hearing families who do not know how to sign
(i.e., those with language deprivation) (Woll, 2013). This means that
the auditory modality used in the household is not accessible to many
deaf children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school. First language deprivation or
delay has been found to have negative effects on certain cognitive and
linguistic measures among late signing children. To name a few, delayed
exposure to a first functional language is reported to negatively affect
deaf children's morphosyntactic production and processing (Boudreault
and Mayberry, 2006; Cheng and Mayberry, 2021; Cormier et al., 2012;
Karadöller et al., 2017; Kayabaşı and Gökgöz, 2022; Newport, 1990) and
narrative abilities (e.g., Becker, 2009). In our overall research, we
are concerned with the effects of this late exposure in the ultimate
adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information (Chafe,
1976). The Givenness Hierarchy is akin to other previously proposed
models of reference control within the topicality and saliency paradigm
such as the Accessibility Hierarchy (Ariel, 1990), Topic Continuity
(Givon, 1983), and Familiarity Scale (Prince, 1981). Gundel et al.
(1993) argued that the distribution of forms is not random but abides by
the Gricean Maxim of Quantity (Grice, 1975). That is to say, individuals
must make the most informative and succinct choice possible to retain
economy of form. When multiple expressions are available for the same
referent as in an utterance like (\ref{ex:1}), phonetically simplest
form is selected with the least redundant information are selected based
on an economy principle. In the utterance, REs for ``Eddie'' are given
in bold. All three REs (nominal, pronominal, and null marking) refer to
the same entity; however, the quantity of marking decreases as the
referent's accessibility increases. On this ground, for highly
accessible referents (i.e., those that are highly active in the mind),
linguistic forms with the least phonetic content are selected for the
least amount of processing effort.

@@

It is also possible to categorize REs into three according to their
discourse status (Gullberg, 2006). A referent can be introduced or
mentioned for the first time, maintained, that is, continued across two
or more clauses, or re-introduced into the discourse following a topic
shift. As a result, interlocutors mark some REs differently than others
based on the principles of reference accessibility. For spoken
languages, this observation is well established (Ahrenholz, 2005; Ariel,
1990; Azar and Özyürek, 2015; Chafe, 1976; Debreslioska et al., 2013;
Givon, 1983; Gundel et al., 1993; Hickmann et al., 1996). These research
indicate that the introduction and re-introduction of a referent are
often accomplished by full noun phrases which can be commonly modified
by an adjective or demonstrative. Maintenance, on the other hand, is
typically marked with pronominals and null markers since the continued
referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives. Building on Gundel et al.
(1993)'s work, Swabey (2002) demonstrated that, like other spoken
languages such as Russian, Chinese, and Japanese, American Sign Language
(ASL) signers use bare nouns for introduced and re-introduced discourse
contexts whereas null marking is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and were thus economic in their reference tracking.
Similar findings have been reported for German Sign Language (Deutsche
Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British Sign
Language (BSL) (Morgan, 2005). It is important to note, however, that
these research either collapsed native and late signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with language deprivation) in terms
of narrative competence.

@@

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions (Lillo-Martin and Klima, 1990).
In addition, the pointing sign is used to referentially anchor a
referent to a location in signing space. For instance, when the signer
introduces \textsc{elvan} for the first time, s/he can assign her to the
right. This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``a'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``b'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represents the referential locus assigned to the referents
in (\ref{ex:3}).

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). Classifiers denote sets of objects based on semantic
criteria such as size, shape, body-part, texture and sometimes function.
Usually, the verbal classifier and the cross-referenced noun co-occur in
the same clause when the argument is introduced into discourse. The
classifier can track the argument on its own on subsequent occurrences.
In sign languages, a verbal classifier morpheme is encoded by a
\emph{handshape} that is articulated phonologically co-temporally with a
movement, which encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). These two infused morphemes form a complex
predicate. Some classifier handshapes and the set of objects that they
denote are below. Some classifier handshapes and the set of objects that
they denote are presented below in (\ref{ex:4}).

@@

The linguistic inventory for discourse referents can include nominal
constructions (bare nouns, fingerspelled nouns, and modified nouns),
null markers (plain verb, agreement verb, and constructed action),
classifier predicates (whole entity classifiers, body part classifiers,
handling classifiers, and extension classifiers), and overt pronouns
(pronominal index or IX) (Frederiksen and Mayberry, 2016). This interim
section describes such constructions attested for sign languages under
four main headings (nominals, pronouns, classifiers, and null marking).

@@

In TİD and many other sign languages, it is possible to use bare noun
phrases as in (\ref{ex:6a}) as well as fingerspelled nouns as in
(\ref{ex:6c}). Fingerspelling refers to the use of the manual alphabet
which is derived or borrowed from a surrounding written language (Kubuş,
2008). Nouns can also be modified by classifiers as in (\ref{ex:6b})
(Frederiksen and Mayberry, 2016) or by a prenominal or postnominal IX
functioning as a determiner or a demonstrative as in (\ref{ex:6d})
(Maclaughlin, 1997).

@@

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse (Emmorey, 1996). The same IX function has also been identified
for TİD (Sevinç, 2006). The example in (\ref{ex:7}) illustrates a
spatially-anchored pronominal reference for topic re-introduction. These
signs are thus rarely used for the first mention of an entity (Perniss
and Özyürek, 2015). Other interpretations consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns (Koulidobrova and Lillo-Martin, 2016). It is important to
mention here that in some spoken languages (e.g., German and Dutch),
demonstratives can also refer to persons (Bosch and Umbach, 2007) and
behave similarly to personal pronouns. However, for the purposes of the
present study, we will refer to IX constructions as pronominal IX and
particularly examine their non-deictic or personal pronoun realizations
as a possible discourse strategy to mainly mark topic shift among native
and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also common and found in TİD (see Dikyuva
et al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL) occur with transitive verbs
and depict how an object is handled or manipulated by agent. On this
ground, an HCL\footnote{As discussed before, an HCL is transitive in
  nature and depicts the handling action performed by ananimate agent.
  This is why most research tend to consider HCLs as an extension to
  constructed action (Cormier et al., 2013; Dudis, 2004).} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

@@

Parallel to spoken languages, many sign languages have been reported to
allow null subjects and/or objects and make use of null arguments when
referring to highly accessible referents which are still active in the
mental representation of the addressee. In signed discourse, this can be
achieved by employing certain verbal constructions or constructed action
(CA). The latter construction exemplified in (\ref{ex:9a}) refers to
cases in which the signers engage with the space in front of themselves
as if they are interacting with the objects and characters in the story.
This is often achieved with the signer taking on the role of a referent
in the narrative and performing the actions (Metzger, 1995). CA is often
marked by certain non-manual cues such as the facial expression and the
breaking of the eye contact. This is to say, the narrator's head, face,
torso, arms, and as an optional marker, hands represent those of the
associated referent(s) (Smith and Cormier, 2014). As for the verbal
constructions, to identify the predicate types in ASL, Padden (1986)
suggested a triadic categorization of verbs in sign languages: plain
verbs, agreement verbs, and spatial verbs. According to this
classification, plain verbs (\ref{ex:9b}) do not agree with the subject
or the object. This means that they are not inflected for number or
person. This contrasts with agreeing verbs (\ref{ex:9c}) the starting or
the ending point of which can be modified to specific loci for person
and number agreement while spatial verbs (as in \textsc{ix book put} `I
placed the book there') agree with locative arguments but like plain
verbs, they are not inflected for number or person. TİD too licenses
null arguments. Following this classification, Sevinç (2006) argued that
although it is common in TİD to drop arguments, there were certain
limitations on the null marking of the arguments for agreement and plain
verbs.

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

>>>>

\documentclass[review]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

@@

\usepackage{lineno} % add

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{floatrow}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{float}
\floatplacement{figure}{H}
\floatplacement{table}{H}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{times}
\usepackage{smartdiagram}
\usepackage{ragged2e}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{./sty/langsci-gb4e}
\usepackage{caption}
\usepackage{multicol}
\setlength{\columnsep}{-3cm}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\begin{frontmatter}

  \title{Reference Tracking Strategies of Native and Late Signers in
Turkish Sign Language}
      \cortext[cor1]{Corresponding author}

  \begin{abstract}
  We investigated the reference tracking strategies among native and
  late adult deaf signers of Turkish Sign Language. Consistent with the
  theories of saliency and referential accessibility, regardless of
  their acquisiton groups, signers mainly used nominals and extension
  classifiers for introducing referents. To maintain a referent in the
  previous clause, signers chose null marking (e.g., constructed action,
  agreement and plain verbs). For topic shifts or re-introduced
  contexts, nominals and pronouns were chiefly favored although we
  observed very little pronominal use. As for age of acquisition
  effects, we only report limited over-redundancy for late signers who
  used null marking less compared to native signers especially for
  introduced and maintained contexts. Given our limited support for
  redundancy by late signers, we conclude that native signers might
  utilize the spatial affordances of the visual-spatial modality better
  than late signers only to a certain extent.
  \end{abstract}
    \begin{keyword}
    Referential accessibility, signed narratives, pragmatic competence,
delayed language acquisition, Turkish Sign Language \sep 
    Referential accessibility, signed narratives, pragmatic competence,
delayed language acquisition, Turkish Sign Language
  \end{keyword}

 \end{frontmatter}

@@

Narratives are extensive discourses that construct an important part of
human communication. In order to create a systematic and coherent
organization of experiences, narrators must choose appropriate
linguistic forms to refer to entities
\citep{ariel1990, givon1983, gundel1993}: Certain forms such as nominal
expressions are commonly used for the first mention of a referent in a
discourse, and other forms such as pronouns and null marking may track
previously mentioned entities which could either be maintained across
multiple clauses or re-introduced back into the discourse. The selection
of these referring expressions (henceforth REs) is contingent on
cognitive abilities, such that individuals must rely on the activation
of the working memory to keep in mind previously referred expressions
and long strings of information \citep{morgan2005, bamberg1997}.
Pragmatic abilities too play a pivotal role in the tracking of
referents. The form with the least amount of redundant information must
be selected based on the accessibility levels of the referent (i.e., how
active a referent is in the addressee's mind given a point in discourse)
\citep{ahn2019, ariel1990}. Because of these cognitive and pragmatic
necessities, it is not surprising to observe that producing a coherent
narrative is one of the latest milestones in language acquisition among
children \citep{hickmann1996}. As can be expected, there is a plethora
of research examining how narrative competence is achieved in both
monolingual and bilingual acquisition contexts among children and
adults. Although narrative competence and referent selection have been
studied extensively for typically developing speakers
\citep{aksu-koc2015, hickmann1996, williams1988}, deaf individuals who
acquired a sign language in atypical learning contexts (i.e., deaf
signers with linguistic deprivation) have remained mostly understudied
with an exception of handful of studies
\citep{cormier2013, gur2018, becker2009}. Nevertheless, it is paramount
that we investigate and describe the effects of language deprivation
among deaf adults on narrative competence more in depth. This will help
us understand whether the loss of systematic and frequent language input
in the early years of life affect how deaf signers produce a narrative
in their adulthood. Given this, the present study will focus on deaf
adult signers with language deprivation and examine their reference
tracking strategies in signed narratives.

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 \citep[see][ for a brief review of the historical development
of the language]{nuhbalaoglu2018}. Deaf or hearing children acquiring a
SL within a family where there is one or more deaf caregivers display
the same developmental milestones as hearing children acquiring a spoken
language in a hearing family
\citep{chamberlain_language_1999, meier1990}. However, reports indicate
that over 90\% of all deaf individuals are born into hearing families
who do not know how to sign (i.e., those with language deprivation)
\citep{woll2013}. This means that the auditory modality used in the
household is not accessible to many deaf children. That is why the age
of acquisition of a first functional language varies greatly among deaf
children, with many of whom being deprived of their first languages
typically until after age 7 when they start formal education in a deaf
school. First language deprivation or delay has been found to have
negative effects on certain cognitive and linguistic measures among late
signing children. To name a few, delayed exposure to a first functional
language is reported to negatively affect deaf children's
morphosyntactic production and processing
\citep{kayabasi2020, boudreault2006, newport1990, cheng2021, karadoller2017, cormier2012}
and narrative abilities \citep[e.g.,][]{becker2009}. In our overall
research, we are concerned with the effects of this late exposure in the
ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy \citep{gundel1993}.
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information
\citep{chafe1976}. The Givenness Hierarchy is akin to other previously
proposed models of reference control within the topicality and saliency
paradigm such as the Accessibility Hierarchy \citep{ariel1990}, Topic
Continuity \citep{givon1983}, and Familiarity Scale \citep{prince1981}.
\citet{gundel1993} argued that the distribution of forms is not random
but abides by the Gricean Maxim of Quantity \citep{grice1975}. That is
to say, individuals must make the most informative and succinct choice
possible to retain economy of form. When multiple expressions are
available for the same referent as in an utterance like (\ref{ex:1}),
phonetically simplest form is selected with the least redundant
information are selected based on an economy principle. In the
utterance, REs for ``Eddie'' are given in bold. All three REs (nominal,
pronominal, and null marking) refer to the same entity; however, the
quantity of marking decreases as the referent's accessibility increases.
On this ground, for highly accessible referents (i.e., those that are
highly active in the mind), linguistic forms with the least phonetic
content are selected for the least amount of processing effort.

@@

It is also possible to categorize REs into three according to their
discourse status \citep{gullberg2006}. A referent can be introduced or
mentioned for the first time, maintained, that is, continued across two
or more clauses, or re-introduced into the discourse following a topic
shift. As a result, interlocutors mark some REs differently than others
based on the principles of reference accessibility. For spoken
languages, this observation is well established
\citep{ariel1990, givon1983, gundel1993, debreslioska2013, ahrenholz2005, chafe1976, hickmann1996, azar2015}.
These research indicate that the introduction and re-introduction of a
referent are often accomplished by full noun phrases which can be
commonly modified by an adjective or demonstrative. Maintenance, on the
other hand, is typically marked with pronominals and null markers since
the continued referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives. Building on
\citet{gundel1993}'s work, \citet{swabey2002} demonstrated that, like
other spoken languages such as Russian, Chinese, and Japanese, American
Sign Language (ASL) signers use bare nouns for introduced and
re-introduced discourse contexts whereas null marking is used for
referents which are ``in-focus'' or very accessible to the addressee.
Overt REs like pronouns or nouns are disprefered for such referents.
\citet{swabey2002} concluded that ``native'' signers in her study abided
by the Gricean maxim of quantity and were thus economic in their
reference tracking. Similar findings have been reported for German Sign
Language (Deutsche Gebärdensprache - DGS) \citep{perniss2015} and
British Sign Language (BSL) \citep{morgan2005}. It is important to note,
however, that these research either collapsed native and late signers
into a single group and labeled them as ``native-like'' or explicitly
compared native deaf signers with late deaf signers (with language
deprivation) in terms of narrative competence.

@@

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions \citep{lillo-martin1990}. In
addition, the pointing sign is used to referentially anchor a referent
to a location in signing space. For instance, when the signer introduces
\textsc{elvan} for the first time, s/he can assign her to the right.
This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``a'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``b'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represents the referential locus assigned to the referents
in (\ref{ex:3}).

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
\citep{aikhenvald2017}. Classifiers denote sets of objects based on
semantic criteria such as size, shape, body-part, texture and sometimes
function. Usually, the verbal classifier and the cross-referenced noun
co-occur in the same clause when the argument is introduced into
discourse. The classifier can track the argument on its own on
subsequent occurrences. In sign languages, a verbal classifier morpheme
is encoded by a \emph{handshape} that is articulated phonologically
co-temporally with a movement, which encodes the verbal meaning
\citep{supalla1990, benedicto2004}. These two infused morphemes form a
complex predicate. Some classifier handshapes and the set of objects
that they denote are below. Some classifier handshapes and the set of
objects that they denote are presented below in (\ref{ex:4}).

@@

The linguistic inventory for discourse referents can include nominal
constructions (bare nouns, fingerspelled nouns, and modified nouns),
null markers (plain verb, agreement verb, and constructed action),
classifier predicates (whole entity classifiers, body part classifiers,
handling classifiers, and extension classifiers), and overt pronouns
(pronominal index or IX) \citep{frederiksen2016}. This interim section
describes such constructions attested for sign languages under four main
headings (nominals, pronouns, classifiers, and null marking).

@@

In TİD and many other sign languages, it is possible to use bare noun
phrases as in (\ref{ex:6a}) as well as fingerspelled nouns as in
(\ref{ex:6c}). Fingerspelling refers to the use of the manual alphabet
which is derived or borrowed from a surrounding written language
\citep{kubus2008}. Nouns can also be modified by classifiers as in
(\ref{ex:6b}) \citep{frederiksen2016} or by a prenominal or postnominal
IX functioning as a determiner or a demonstrative as in (\ref{ex:6d})
\citep{maclaughlin1997}.

@@

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse \citep{emmorey1996}. The same IX function has also been
identified for TİD \citep{sevinc2006}. The example in (\ref{ex:7})
illustrates a spatially-anchored pronominal reference for topic
re-introduction. These signs are thus rarely used for the first mention
of an entity \citep{perniss2015}. Other interpretations consider
pointing signs without noun phrases as examples of demonstratives and
not personal pronouns \citep{koulidobrova2016}. It is important to
mention here that in some spoken languages (e.g., German and Dutch),
demonstratives can also refer to persons \citep{bosch2007} and behave
similarly to personal pronouns. However, for the purposes of the present
study, we will refer to IX constructions as pronominal IX and
particularly examine their non-deictic or personal pronoun realizations
as a possible discourse strategy to mainly mark topic shift among native
and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages \citep{supalla1986}. These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape \citep{talmy2003}. As in most sign languages hitherto studied,
classifier constructions are also common and found in TİD \citep[see][
for a review]{dikyuva2015}. Following the classification of
\citet{engberg-pedersen1993}, we distinguish between four types of
classifier constructions. Whole entity classifiers (WECL) represent the
entirety of any referent whereas body part classifiers (BPCL) only
reflect a specific part of an animate entity (e.g., limbs, head, tongue,
teeth, and so forth). Handling classifiers (HCL) occur with transitive
verbs and depict how an object is handled or manipulated by agent. On
this ground, an HCL\footnote{As discussed before, an HCL is transitive
  in nature and depicts the handling action performed by ananimate
  agent. This is why most research tend to consider HCLs as an extension
  to constructed action \citep{dudis2004, cormier2013}.} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

@@

Parallel to spoken languages, many sign languages have been reported to
allow null subjects and/or objects and make use of null arguments when
referring to highly accessible referents which are still active in the
mental representation of the addressee. In signed discourse, this can be
achieved by employing certain verbal constructions or constructed action
(CA). The latter construction exemplified in (\ref{ex:9a}) refers to
cases in which the signers engage with the space in front of themselves
as if they are interacting with the objects and characters in the story.
This is often achieved with the signer taking on the role of a referent
in the narrative and performing the actions \citep{metzger1995}. CA is
often marked by certain non-manual cues such as the facial expression
and the breaking of the eye contact. This is to say, the narrator's
head, face, torso, arms, and as an optional marker, hands represent
those of the associated referent(s) \citep{smith2014}. As for the verbal
constructions, to identify the predicate types in ASL,
\citet{padden1986} suggested a triadic categorization of verbs in sign
languages: plain verbs, agreement verbs, and spatial verbs. According to
this classification, plain verbs (\ref{ex:9b}) do not agree with the
subject or the object. This means that they are not inflected for number
or person. This contrasts with agreeing verbs (\ref{ex:9c}) the starting
or the ending point of which can be modified to specific loci for person
and number agreement while spatial verbs (as in \textsc{ix book put} `I
placed the book there') agree with locative arguments but like plain
verbs, they are not inflected for number or person. TİD too licenses
null arguments. Following this classification, \citet{sevinc2006} argued
that although it is common in TİD to drop arguments, there were certain
limitations on the null marking of the arguments for agreement and plain
verbs.

@@

\renewcommand\refname{References}
\bibliography{./bibliography/references.bib}
